Inferring stellar rotation periods using Gaussian processes

### 2016-12-20 documents/ms.tex

Sampling the posterior described above presents several challenges:
\begin{itemize}
\item Evaluating $\mathcal L$ for an entire \Kepler\ lightcurve 
($\sim$40,000 points) takes about $\sim$5\,s--- too computationally expensive 
to perform inference on large numbers of light curves\footnote{All computational
times cited in this section are based on evaluations on a 
single core of a 2015 Macbook Pro, 3.1 GHz Intel Core i7.}.  The matrix operations
necessary to evaluate the GP likelihood (using the fast matrix solver 
HODLR \citep{Ambikasaran2014}, implemented in the {\tt george} \citep{George} 
python package) scale as $N\ln(N)^2$, where $N$ is
the number of data points in the light curve.

\item The flexibility of this GP model allows for insidious kinds of 
posterior multimodality and "over-fitting"--like behavior.  For example, 
if $l$ is small,  the non-periodic factor in the covariance kernel may dominate, 
allowing for a good fit to the data without requiring any 
periodic covariance structure---even if
clear periodic structure is present.  Additionally, if $\Gamma$ is too large,
then [explain in more technical terms that "the GP gets way too wiggly and
overfits the data without caring about the periodicity." ]

@@

We address the first of these difficulties in two distinct ways: subsampling
the data and splitting the light curve into independent sections.  
To subsample, we randomly select 1/30th of the points in the full light curve (an 
average of $\sim$1.5 points per day).  This decreases the likelihood evaluation 
time by a factor of about 50, down to about 100\,ms.
We then split the light curve into equal-sized
chunks containing approximately 300 points per section (corresponding to about
200 days), and evaluate the log-likelihood as the sum of the log-likelihoods
of the individual sections (all using the same parameters $\theta$).  This reduces
computation time because the section-based likelihood evaluation scales as $mn\ln(n)^2$, where $n$ is the number of data points per section and $m$ is the number of sections.
This method further reduces computation time for a typical light curve (subsampled
by a factor of 30) by about a factor of two, to about 50\,ms.

In order to manage the extreme flexibilty of the GP model and focus on 
reliably retrieving the period parameter, we also impose priors
on the non-period GP parameters that we 
learned during the development of this method.  
Beginning with very broad priors (uniform in log space between 
-20 and 20) on all the parameters, we first explored what the 
typical values of the non-period parameters tended to be when the 
injected periods were successfully recovered.  We also experimented with
constraining the allowed ranges of the parameters, after discovering 
that some regions of parameter space (such as large values of $A$ and $\Gamma$
and small values of $l$) tended to allow fits that ignored the desired
periodicity.  The priors we used for this work are listed in
table \ref{tab:priors}.  
% We note that our default prior for period is 
% log-uniform between the bounds, but that we were also able to 
% further improve our results by using an ACF-inspired prior.

@@

Figure \ref{fig:compare_mcmc} summarizes the results of this MCMC fitting procedure
compared to the injected ``true'' stellar rotation periods.

@@

The marginal posterior distributions of the QP kernel hyperparameters for the
example simulated light curve with in figure \ref{fig:demo_lc}, are shown in
figure \ref{fig:gp_posteriors}.

>>>>

Sampling the posterior described above presents several challenges:
\begin{itemize}
\item Evaluating $\mathcal L$ for an entire \Kepler\ lightcurve 
($\sim$40,000 points) takes about $\sim$5\,s--- too computationally expensive 
to perform inference on large numbers of light curves\footnote{All computational
times cited in this section are based on evaluations on a 
single core of a 2015 Macbook Pro, 3.1 GHz Intel Core i7.}.  The matrix operations
necessary to evaluate the GP likelihood\footnote{These operations use 
the fast matrix solver 
HODLR \citep{Ambikasaran2014}, implemented in the {\tt george} \citep{George} 
python package.} scale as $N\ln(N)^2$, where $N$ is
the number of data points in the light curve.

\item The flexibility of this GP model allows for insidious kinds of 
posterior multimodality and ``over-fitting''--like behavior.  For example, 
if $l$ is small,  the non-periodic factor in the covariance kernel may dominate, 
allowing for a good fit to the data without requiring any 
periodic covariance structure---even if
clear periodic structure is present.  Additionally, if $\Gamma$ is too large,
then [explain in more technical terms that ``the GP gets way too wiggly and
overfits the data without caring about the periodicity.'' ]

@@

We address the difficulty of computational speed in two distinct ways: subsampling
the data and splitting the light curve into independent sections.  
To subsample, we randomly select 1/30th of the points in the full light curve (an 
average of $\sim$1.5 points per day).  This decreases the likelihood evaluation 
time by a factor of about 50, down to about 100\,ms.
We then split the light curve into equal-sized
chunks containing approximately 300 points per section (corresponding to about
200 days), and evaluate the log-likelihood as the sum of the log-likelihoods
of the individual sections (all using the same parameters $\theta$).  This reduces
computation time because the section-based likelihood evaluation scales as $mn\ln(n)^2$, where $n$ is the number of data points per section and $m$ is the number of sections.
This method further reduces computation time for a typical light curve (subsampled
by a factor of 30) by about a factor of two, to about 50\,ms.

In order to manage the extreme flexibilty of the GP model and focus on 
reliably retrieving the period parameter, we also impose priors
on the non-period GP parameters that we 
learned during the development of this method.  
Beginning with very broad priors (uniform in log space between 
-20 and 20) on all the parameters, we first explored what the 
typical values of the non-period parameters tended to be when the 
injected periods were successfully recovered.  We also experimented with
constraining the allowed ranges of the parameters, after discovering 
that some regions of parameter space (such as large values of $A$ and $\Gamma$
and small values of $l$) tended to allow fits that ignored the desired
periodicity.  The priors we used for this work are listed in
table \ref{tab:priors}.  We note that we do not have any particular
quantitative justification for the use of these priors,
except that they produce good results for this particular test set of 
light curves.  We thus caution that other data sets might have different
noise properties, potentially requiring modifications to these priors.
% We note that our default prior for period is 
% log-uniform between the bounds, but that we were also able to 
% further improve our results by using an ACF-inspired prior.

@@

@@

Figure \ref{fig:compare_mcmc} summarizes the results of this MCMC fitting procedure
compared to the injected ``true'' stellar rotation periods. 
The marginal posterior distributions of the QP kernel hyperparameters for the
example simulated light curve with in figure \ref{fig:demo_lc}, are shown in
figure \ref{fig:gp_posteriors}.
